---
title: "Breakpoint Analysis of Trends in River Exports"
author: "Nick J Lyon"
output: pdf_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Breakpoint Analysis via `SiZer`

`SiZer` is an R package that identifies "significant zero crossings of the derivative." In simpler terms, `SiZer` identifies points on a trendline where the slope changes. Identifying these slope changes can be helpful in 'splitting' a line (i.e., identifying breakpoints) and comparing various metrics of the resulting line segments.

## Preparing to Use `SiZer`

This report is meant to demonstrate one possible workflow for identifying these breakpoints and then using them to analyze chunks of a trendline separately. To run this, you will need the following R packages.

```{r libraries, message = F, warning = F}
library(broom); library(cowplot); library(SiZer); library(tidyverse)
```

To use this version of the workflow, you need a dataframe containing (at least) a column of a continuous X variable (i.e., explanatory/independent variable) and a column of a continuous Y variable (i.e., response/dependent variable). The dataset loaded for this example includes data from many sites and the workflow loops across these sites and identifies the breakpoints for each.

```{r load-data, message = F}
# Load data
data <- readr::read_csv(file = file.path("data", "CryoData_forNick_6.29.22.csv"))

# Briefly, check it out
dplyr::glimpse(data)
```

We (at the LNO Scientific Computing Support Team) have designed a suite of helper functions to facilitate this workflow and cut down on the number of necessary lines of code. We'll describe these functions in greater detail as we continue but for now, we can load the R package that includes these functions: `HERON`. `HERON` is a loose acronym for *HE*lpers for *R*iver *O*bservatio*N* given that its functions were developed with hydrological research in mind.

```{r load-helper-fxns-demo, eval = F}
# Install HERON
# install.packages("remotes")
remotes::install_github("lter/HERON")
library(HERON)
```

```{r load-helper-fxns, include = F}
# Get the HERON package
librarian::shelf(lter/HERON)
```

To facilitate iterating across response variables and sites, the script this report is generated from allows the user to pick the response and explanatory variables and assign them as objects. Also, the loop normally iterates across sites but for the purposes of this report we can select a single site here

```{r load-variables}
# Identify response (Y) and explanatory (X) variables
response_var <- "FNYield"
explanatory_var <- "Year"

# Identify desired site
place <- "ALBION"
```

Also, `SiZer` assesses "bandwidths" that define the width of the 'window' used to evaluate slope changes. A very low bandwidth (i.e., a narrow window) will see a single point that deviates from the general trend and will identify a slope change on either side of that point. A higher bandwidth (e.g. a wide window) will only identify the most obvious slope changes and will often miss subtle hills or valleys in a trendline.

Choosing the right bandwidth is a matter of some debate and relies heavily upon exploration of the data and expert intuition so this workflow makes the user select three bandwidths to help guide this necessary exploration.

```{r load-bandwidths}
# Identify the three bandwidths you want to look at specifically
band_low <- 4
band_mid <- 5
band_high <- 8
```

Last preparation step, but because this workflow is a large loop, we must make an empty list to store various products and create a counter to advance when we move from one site to the next.

```{r loop-prep}
# Make an empty list to store all of our extracted information
giant_list <- list()

# Make a counter and set it to 1 (the list will add to it)
j <- 1
```

## Fundamental Workflow - Initial Fit

The loop begins by subsetting the data to only a single site. The rest of the workflow assumes you're only identifying breakpoints in a single trendline so it would certainly error out if you attempted to handle several separate trendlines at once.

```{r subset-data}
# Subset the data to desired site
  data_sub <- data %>%
    dplyr::filter(site == place) %>%
    as.data.frame()
```

Once we've subsetted the data, we can invoke `SiZer` to identify the slope changes across a range of bandwidths. You can see how the subsetted data, response, and explanatory variables are invoked below.

The `h` argument defines which bandwidths to assess and 2 to 10 is a good expert-defined upper and lower boundary. This makes `SiZer` identify slope changes for every bandwidth between 2 and 10. The `degree`, `derv`, and `grid.length` arguments are specified for clarity but are unimportant at this level of discussion. Consult `?SiZer::SiZer` for the package authors' explanation of these arguments.

```{r run-sizer, message = F}
# Invoke the SiZer::SiZer function
  e <- SiZer::SiZer(x = data_sub[[explanatory_var]],
                    y = data_sub[[response_var]],
                    h = c(2, 10), degree = 1,
                    derv = 1, grid.length = 100)
```

Once we have a `SiZer` object, we can plot it to evaluate slope changes across all provided bandwidths. The custom `sizer_plot` function plots this and adds horizontal lines at the user-selected bandwidths.

This plot can be a little dense but here is a brief summary of the key information:
- The x-axis is the same as the explanatory variable as it appears in the data
- Each "row" of pixels represents the assessment of slope changes for a single bandwidth
- **Blue** pixels are a **positive** slope, **red** pixels are a **negative** slope, and **purple** are a **flat** slope
- Gray pixels **do not have a slope assessment**

You can see in the plot below that higher bandwidths (e.g., upper rows of pixels) identify fewer slope changes (i.e., future changes in pixel color when looking within a row from left to right) while lower bandwidths identify more slope changes.

```{r sizer-plot}
# Plot the SiZer object with horizontal lines at bandwidths of interest
  HERON::sizer_plot(sizer_object = e, bandwidth_vec = c(band_low, band_mid, band_high))
```

We then found it useful to do two things with this object:

- First, we identify the slope changes for each of the three specified bandwidths
- Second, we aggregate within a 'column' of pixels and identify the "average" slope changes across all bandwidths. This is somewhat of an abstraction from the data but does avoid any need for an expert judgement call on which bandwidth(s) should be evaluated.

`sizer_slice` and `sizer_aggregate` are custom functions written for these specific purposes by us.

```{r sizer-pull}
# Identify inflection points at three specific bandwidths too
sizer_low <- HERON::sizer_slice(sizer_object = e, bandwidth = band_low)
sizer_mid <- HERON::sizer_slice(sizer_object = e, bandwidth = band_mid)
sizer_high <- HERON::sizer_slice(sizer_object = e, bandwidth = band_high)

# Strip out the aggregated (across all bandwidths) inflection points
sizer_tidy <- HERON::sizer_aggregate(sizer_object = e)
```

## Fundamental Workflow - Identify Slope Changes

Now that we've stripped the relevant parts of the `SiZer` object, we want to take the slope changes identified there and put them back in ur data for downstream use.

```{r sizer-data-connect}
# Identify inflection points
## Aggregate
data_sub_agg <- HERON::id_slope_changes(raw_data = data_sub, 
                                 sizer_data = sizer_tidy,
                                 x = explanatory_var,
                                 y = response_var)
## Low
data_sub_low <- HERON::id_slope_changes(raw_data = data_sub,
                                 sizer_data = sizer_low,
                                 x = explanatory_var,
                                 y = response_var)
## Mid
data_sub_mid <- HERON::id_slope_changes(raw_data = data_sub,
                                 sizer_data = sizer_mid,
                                 x = explanatory_var,
                                 y = response_var)
## High
data_sub_high <- HERON::id_slope_changes(raw_data = data_sub,
                                  sizer_data = sizer_high,
                                  x = explanatory_var,
                                  y = response_var)
```

Now that we migrated the `SiZer` slope changes into the data, we can create exploratory plots of our trendline with the change points that `SiZer` said match changes in slope.

This is essentially a more digestible version of the `SiZer` plot that is red/blue/purple.

```{r sizer-plots-create, echo = F, warning = F, message = F}
# Plot the aggregated inflection points
agg_plot <- HERON::sizer_ggplot(raw_data = data_sub_agg,
                         sizer_data = sizer_tidy,
                         x = explanatory_var, y = response_var,
                         trendline = 'sharp', vline = "none",
                         sharp_colors = c("#bbbbbb", 'orange')) +
  ggtitle(label = "Aggregated Slope Changes")

# Plot the bandwidth-specific plots too!
## Low Bandwidth (h)
low_plot <- HERON::sizer_ggplot(raw_data = data_sub_low,
                         sizer_data = sizer_low,
             x = explanatory_var, y = response_var,
             trendline = 'sharp', vline = "none",
             sharp_colors = c("#bbbbbb", 'orange')) +
  ggtitle(label = paste0("h = ", band_low, " Slope Changes"))
## Mid Bandwidth (h)
mid_plot <- HERON::sizer_ggplot(raw_data = data_sub_mid,
                         sizer_data = sizer_mid,
             x = explanatory_var, y = response_var,
             trendline = 'sharp', vline = "none",
             sharp_colors = c("#bbbbbb", 'orange')) +
  ggtitle(label = paste0("h = ", band_mid, " Slope Changes"))
## High Bandwidth (h)
high_plot <- HERON::sizer_ggplot(raw_data = data_sub_high,
                          sizer_data = sizer_high,
             x = explanatory_var, y = response_var,
             trendline = 'sharp', vline = "none",
             sharp_colors = c("#bbbbbb", 'orange')) +
  ggtitle(label = paste0("h = ", band_high, " Slope Changes"))
```

In these plots, orange lines are either an increasing or deacreasing slope while gray lines are flat slopes. The title indicates which bandwidth is being considered in that panel of the combination plot.

```{r sizer-plots-show, echo = F, warning = F, message = F}
cowplot::plot_grid(agg_plot, low_plot, mid_plot, high_plot,
                   nrow = 2, ncol = 2, labels = "AUTO")
```

The workflow exports these data files (that contain both data and the important findings of the `SiZer` object) for later use but we can ignore that for this report.

```{r export-wrangled-data, eval = F, echo = F}
# Now modify the columns in the provided sizer dataframes
sizer_tidy_export <- sizer_tidy %>%
  # Make everything a character
  dplyr::mutate(dplyr::across(dplyr::everything(), as.character)) %>%
  # Add a column for bandwidth and for site name
  dplyr::mutate(site = place, h_grid = "averaged across bandwidths",
                .before = dplyr::everything() ) %>%
  # Make sure it's a dataframe
  as.data.frame()

# Do the same for the bandwidth specific data
## Low Bandwidth
sizer_low_export <- sizer_low %>%
  dplyr::mutate(dplyr::across(dplyr::everything(), as.character)) %>%
  dplyr::mutate(site = place, .before = dplyr::everything()) %>%
  as.data.frame()
## Mid Bandwidth
sizer_mid_export <- sizer_mid %>%
  dplyr::mutate(dplyr::across(dplyr::everything(), as.character)) %>%
  dplyr::mutate(site = place, .before = dplyr::everything()) %>%
  as.data.frame()
## High Bandwidth
sizer_high_export <- sizer_high %>%
  dplyr::mutate(dplyr::across(dplyr::everything(), as.character)) %>%
  dplyr::mutate(site = place, .before = dplyr::everything()) %>%
  as.data.frame()

# Combine the specific bandwidth objects
complete_export <- dplyr::bind_rows(sizer_low_export,
                                    sizer_mid_export,
                                    sizer_high_export)

# Add these tidied dataframes to our lists
giant_list[[paste0("aggregate_", j)]] <- sizer_tidy_export
giant_list[[paste0("specific_", j)]] <- complete_export
```

## Fundamental Workflow - Analyze Data with Identified Breakpoints

With these data objects that contain both the trendline *and* the slope changes we can move forward with the actual *analysis* part of breakpoint analysis.

We can use our custom `sizer_lm` function to fit a linear regression on each "chunk" of the trendline (i.e., between each of the breakpoints identified above). `sizer_lm` returns the summary statistics of the regression and estimates of slope and y-intercepts.

```{r sizer-lm, warning = F, message = F}
# Extract (1) statistics and (2) estimates from linear models
agg_lm <- HERON::sizer_lm(data = data_sub_agg, x = explanatory_var,
                   y = response_var, group_col = "groups") %>%
  # Then add columns for which bandwidth & which site
  purrr::map(.f = mutate, bandwidth = "aggregate",
             .before = dplyr::everything())
## Low bandwidth
low_lm <- HERON::sizer_lm(data = data_sub_low, x = explanatory_var,
                   y = response_var, group_col = "groups") %>%
  purrr::map(.f = mutate, bandwidth = band_low,
             .before = dplyr::everything())
## Middle bandwidth
mid_lm <- HERON::sizer_lm(data = data_sub_mid, x = explanatory_var,
                   y = response_var, group_col = "groups") %>%
  purrr::map(.f = mutate, bandwidth = band_mid,
             .before = dplyr::everything())
## High bandwidth
high_lm <- HERON::sizer_lm(data = data_sub_high, x = explanatory_var,
                    y = response_var, group_col = "groups") %>%
  purrr::map(.f = mutate, bandwidth = band_high,
             .before = dplyr::everything())
```

In the true workflow we then grab these and put them in our "giant_list" object created at the start of this script. For this report though, that is the end of the workflow for breakpoint analysis!

```{r export-part-1, echo = F, eval = F}
# Form one big list
mega_lm_list <- list(agg_lm, low_lm, mid_lm, high_lm)

# Final dataframe processing for *statistics*
stat_df <- mega_lm_list %>%
  # Extract first list element
  purrr::map(.f = 1) %>%
  # Make all columns characters
  purrr::map(.f = dplyr::mutate, dplyr::across(dplyr::everything(),
                                        as.character)) %>%
  # Add a site column
  purrr::map(.f = mutate, site = place,
             .before = dplyr::everything()) %>%
  # Combine all list elements into a dataframe
  purrr::map_dfr(.f = dplyr::select, dplyr::everything())

# Final dataframe processing for *estimates*
est_df <- mega_lm_list %>%
  purrr::map(.f = 2) %>%
  purrr::map(.f = dplyr::mutate, dplyr::across(dplyr::everything(),
                                               as.character)) %>%
  purrr::map(.f = mutate, site = place,
             .before = dplyr::everything()) %>%
  purrr::map_dfr(.f = dplyr::select, dplyr::everything())

# Add this information to their respective lists
giant_list[[paste0("stats_", j)]] <- stat_df
giant_list[[paste0("estimates_", j)]] <- est_df

# Increase the counter by 1 (for the next iteration of the loop)
j <- j + 1
```

Over the course of the whole loop we then have *many* list elements (four per site!) so we have a chunk of code to unlist and export one dataframe for each of the four list elements across site. This step makes interpretation and comparison much simpler than retaining everything in list format.

```{r export-part-2, eval = F, echo = F}
# Check out what is in our huge list
names(giant_list)

# Now (ironically) we'll use a loop to unlist what the first loop made
for(data_type in c("aggregate", "specific", "stats", "estimates")){
  
  # For each data type...
  list_sub <- giant_list %>%
    # Identify all list elements that contain this type of data
    purrr::keep(.p = stringr::str_detect(string = names(.),
                                         pattern = data_type)) %>%
    # Unlist by selecting all columns of each list element
    purrr::map_dfr(.f = dplyr::select, dplyr::everything())
  
  # Now save the CSV
  write_csv(x = list_sub, na = "",
            file = file.path(export_folder,
                             paste0("_slope-change_", data_type, "_exported.csv")))
  
  # And print a message
  message("Dataframe for ", data_type, " exported.") }
```
